---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    echo: true
---

## Intro

This assignment makes use of data from a personal activity monitoring device.  
The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

    
## Initialisation 

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

First load the needed libraries / packages and set the working directory  
```{r}
    library(ggplot2) # for plotting
    library(dplyr) # for manipulating data 
    library(Hmisc) # for the impute function (had to install this package first in rstudio)
    setwd("~/Documents/Machine Learning/reproducible research/RepData_PeerAssessment1")
```

## Loading and preprocessing the data

```{r}
    activData <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?

```{r}
    activeDays <- activData %>% group_by(as.Date(date)) %>% summarise(totalSteps=sum(steps))
    colnames(activeDays) <- c("date", "totalSteps")
    
    meanSteps <- mean((activeDays$totalSteps), na.rm = TRUE)
    medianSteps <- median((activeDays$totalSteps), na.rm = TRUE)
    
    #histTotalSteps <- 
        ggplot(data=activeDays, aes(x=totalSteps)) + 
            geom_histogram(stat="bin", 
                        col="dark green", 
                        fill="green", 
                        alpha = .2) + 
            labs(title="Histogram for mean total number of steps taken per day", x="Number of steps", y="Freqency") +
            geom_vline(aes(xintercept=meanSteps), color="blue", linetype="dashed", size=1, vjust=0.5) + 
            geom_text(aes(x=meanSteps, label="mean", y=8.5), colour="blue", angle=90, vjust=1.2, text=element_text(size=6)) +
            geom_vline(aes(xintercept=meanSteps), color="red", linetype="dashed", size=0.5) + 
            geom_text(aes(x=meanSteps, label="median", y=8.5), colour="red", angle=90, vjust=2.4, text=element_text(size=6))
    
    #print(histTotalSteps)
    cat("Mean total number of steps: ", meanSteps)
    cat("Median total number of steps: ", medianSteps)

```


## What is the average daily activity pattern?

```{r}
    averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activData$steps), by=list(interval=activData$interval), FUN=mean, na.rm=TRUE)
    
    ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
        geom_line(colour="blue") + 
        xlab("Intervals") +
        ylab("Average daily steps") 
```
  
  
## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)  
Calculation of the missing values in the set: 

```{r}
    missingCount <- sum(is.na(activData$steps))
    cat("Number of missing values: ", missingCount)
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
Create a new dataset with the missing values added.   
The impute function is designed for this task: import / add missing data to a set based on a function, I choose the mean function. 
Impute is from the Hmisc package.   

First create a new set, then add the missing values using the mean parameter with the impute function.   

```{r}
    activDataCompleted <- activData
    activDataCompleted$steps <- impute(activData$steps, fun=mean)
```

## What is mean total number of steps taken per day taken into account the imputed values ?  
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
    activeDays <- activDataCompleted %>% group_by(as.Date(date)) %>% summarise(totalSteps=sum(steps))
    colnames(activeDays) <- c("date", "totalSteps")
    
    meanSteps <- mean((activeDays$totalSteps), na.rm = TRUE)
    medianSteps <- median((activeDays$totalSteps), na.rm = TRUE)
    
    # histTotalSteps <- 
        ggplot(data=activeDays, aes(x=totalSteps)) + 
            geom_histogram(stat="bin", 
                        col="dark green", 
                        fill="green", 
                        alpha = .2) + 
            labs(title="Histogram for mean total number of steps taken per day with imouted data", x="Number of steps", y="Freqency") +
            geom_vline(aes(xintercept=meanSteps), color="blue", linetype="dashed", size=1, vjust=0.5) + 
            geom_text(aes(x=meanSteps, label="mean", y=8.5), colour="blue", angle=90, vjust=1.2, text=element_text(size=6)) +
            geom_vline(aes(xintercept=meanSteps), color="red", linetype="dashed", size=0.5) + 
            geom_text(aes(x=meanSteps, label="median", y=8.5), colour="red", angle=90, vjust=2.4, text=element_text(size=6))
    
    # print(histTotalSteps)
    cat("Mean total number of steps incl imputed data: ", meanSteps)
    cat("Median total number of steps incl imputed data: ", medianSteps)

```



## Are there differences in activity patterns between weekdays and weekends?  

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.  
Using the wday function, returns 0–6 day of the week, starting on Sunday. I used the days of week with 0 and 6 as weekend. And 1 to 5 as work week. 


```{r}
    activDataCompleted <- activDataCompleted %>% mutate(dateType = ifelse(as.POSIXlt(activDataCompleted$date)$wday %in% c(1,5), 'weekend', 'weekday'))
```
  
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.  

```{r}
    activDataCompleted <- aggregate(steps ~ interval + dateType, data=activDataCompleted, mean)
    ggplot(activDataCompleted, aes(interval, steps)) + 
        geom_line(colour="blue") + 
        facet_grid(dateType ~ .) +
        xlab("Intervals") + 
        ylab("Mean number of steps")
```