---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    echo: true
---

## Intro

This assignment makes use of data from a personal activity monitoring device.  
The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.  

    
## Initialisation 

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

First load the needed libraries / packages and set the working directory  
```{r}
    library(ggplot2) # for plotting
    library(dplyr) # for manipulating data 
    setwd("~/Documents/Machine Learning/reproducible research/RepData_PeerAssessment1")
```

## Loading and preprocessing the data

```{r}
    activData <- read.csv('activity.csv')
```

## What is mean total number of steps taken per day?

```{r}
    activeDays <- activData %>% group_by(as.Date(date)) %>% summarise(totalSteps=sum(steps))
    colnames(activeDays) <- c("date", "totalSteps")
    
    meanSteps <- mean((activeDays$totalSteps), na.rm = TRUE)
    medianSteps <- median((activeDays$totalSteps), na.rm = TRUE)
    
    histTotalSteps <- ggplot(data=activeDays, aes(x=totalSteps)) + 
            geom_histogram(stat="bin", 
                        col="dark green", 
                        fill="green", 
                        alpha = .2) + 
            labs(title="Histogram for mean total number of steps taken per day", x="Number of steps", y="Freqency") +
            geom_vline(aes(xintercept=meanSteps), color="blue", linetype="dashed", size=1, vjust=0.5) + 
            geom_text(aes(x=meanSteps, label="mean", y=8.5), colour="blue", angle=90, vjust=1.2, text=element_text(size=6)) +
            geom_vline(aes(xintercept=meanSteps), color="red", linetype="dashed", size=0.5) + 
            geom_text(aes(x=meanSteps, label="median", y=8.5), colour="red", angle=90, vjust=2.4, text=element_text(size=6))
    
    print(histTotalSteps)
    cat("Mean total number of steps: ", meanSteps)
    cat("Median total number of steps: ", medianSteps)

```


## What is the average daily activity pattern?

```{r}
    
```


## Imputing missing values

```{r}
    
```


## Are there differences in activity patterns between weekdays and weekends?

```{r}
    
```

